\begin{multicols*}{2}

\section{A}

\section{I}

\begin{entrylist}
    \begin{dictentry}{í\ppa{}ạyạ}{
        iamb., underlyingly \enl{\suph{}í\ppa{}ạya}, from Old Elv. \proto{hi\lgth popa\lgth hi\glotstop a\lgth} `almond flour' from \proto{\nameref{OE:hiipopaa}} `grind, crush' + \proto{\nameref{OE:hi7aa}} `almond'

        irr. singulative: \enl{yo\ppa\ppa ạyạ} `a grain of flour'
    }\label{enl:iphaya_HLL}
        \dictdef*[n., coll.]{almond flour, almond meal}
    \end{dictentry}
\end{entrylist}

\begin{entrylist}
    \begin{dictentry}[il\toneH jũn\toneM]{ílyon}{
        dacty., from Old Elv. \proto{\nameref{OE:hiilino}} `speech'
    }\label{enl:ilyon_HM}
        \dictdef*[n.]{%
            utterance, statement, sentence
        }%
    \end{dictentry}
    \begin{entrysublist}
        \begin{dictentry}{ẹn ílyon}{}\label{enl:en_L_ilyon_HM}
            \dictdef*[v.]{%
                to be sapient, to be intelligent, to be mentally aware, to have one's faculties (lit., to have speech)
            }%
        \end{dictentry}
    \end{entrysublist}
\end{entrylist}

\section{E}

\begin{entrylist}
    \begin{dictentry}{ẹn}{
        troch., from Old Elv. \proto{\nameref{OE:7iina}}
    }\label{enl:en_L}
        \dictdef*{
            there is, to exist, existential particle
        }
    \end{dictentry}
    \begin{dictentry}{ẹnị}{
        particle, from Old Elv. \proto{\glotstop i\lgth na\glotstop i\lgth ho} `there is no' from \proto{\nameref{OE:7iina}} `there is' + \proto{\nameref{OE:-7iiho}} \textsc{neg}
    }\label{enl:eni_LL}
        \dictdef*{
            existential negation particle
        }
    \end{dictentry}
    \begin{dictentry}[ẽ\engma\toneH\vell a\toneL lẽn\toneM]{énłạlen}{
        troch., from Old Elv. \proto{hi\lgth na\glotstop ala\lgth hili\lgth no} `World-speech' from \proto{\nameref{OE:hiina7alaa}} `the World' + \proto{\nameref{OE:hiilino}} `speech'
        }
        \label{enl:enLalen_HLM}
        \dictdef*[n., coll.]{%
            this language, En\l alen
        }%
    \end{dictentry}
\end{entrylist}

\section{O}

\begin{entrylist}
    \begin{dictentry}{ófị}{
        particle, from Old Elv. \proto{ho\lgth fi\glotstop i\lgth ho} `is not at' from \proto{\nameref{OE:hoofi}} `to be located' + \proto{\nameref{OE:-7iiho}} \textsc{neg}
    }\label{enl:ofi_HL}
        \dictdef*{
            imperfective negation particle
        }
    \end{dictentry}
\end{entrylist}

\section{R}

\section{Y}

\section{W}

\section{P}

\begin{entrylist}
    \begin{dictentry}{pol}{
        troch., from Old Elv. \proto{\nameref{OE:poola}} `two'
    }\label{enl:pol_M}
        \dictdef*[num.]{
            two
        }
    \end{dictentry}
\end{entrylist}

\section{T}

\begin{entrylist}
    \begin{dictentry}{tóyó}{
        particle, from Old Elv. \proto{to\lgth\glotstop iho\lgth} `did not' from \proto{\nameref{OE:too}} `to do, to make' + \proto{\nameref{OE:-7iiho}} \textsc{neg}
    }\label{enl:toyo_HH}
        \dictdef*{
            perfective negation particle
        }
    \end{dictentry}
\end{entrylist}

\section{K}

\section{\Ppa}

\section{\Tta}


\begin{entrylist}
    \begin{dictentry}{\tta ẹmo}{
        iamb., from Old Elv. \proto{\nameref{OE:to7iimomoo}} `eyes'
    }\label{enl:themo_LM}
        \dictdef*[n., coll.]{eyes}
    \end{dictentry}
\end{entrylist}

\section{\Kka}

\section{F}

\section{S}

\begin{entrylist}
    \begin{dictentry}{séngạ}{
        iamb., from Old Elv. \proto{\nameref{OE:siini7aa}} `three'
    }\label{enl:senga_HL}
        \dictdef*[num.]{
            three
        }
    \end{dictentry}
\end{entrylist}

\section{M}

\begin{entrylist}
    \begin{dictentry}{matí}{
        dacty., underlyingly \enl{matí\suphi{}}, from Old Elv. \proto{\nameref{OE:pamaa7ihi}} `person, elf'
    }\label{enl:mati_MH}
        \dictdef*[n., sg.]{person, elf}
    \end{dictentry}
\end{entrylist}

\section{N}

\section{\Engga}

\section{L}

\section{Ł}

\end{multicols*}